Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-07-29 tex/draft.tex
Edit analyses and edit/expand viability

\title{Naturalness and Long-lived Beyond the Standard Model Particles}
\author{Bryan Cardwell}

@@

\section{Supersymmetry}
    \fxnote{seems like I should mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

@@

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that stop completely before decaying.

    Displaced vertex and disappearing track analyses both look for particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles within the inner volume \fxnote{specify} of the detector, leaving tracks that point back to a vertex some measurable distance from the nominal interaction point (IP). Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover a wide range of lifetimes and topologies for LLPs that decay on detector timescales\fxnote{dangerous as hell. Be really good at giving a quantitative explanation of this sentence}.

    Slow and stopped particle searches, on the other hand, are sensitive to detector-stable\fxnote{define?} LLPs that are far more massive\fxnote{be able to quantify} than any stable SM particle. The lower velocity of these high-mass particles shows up as increased ionization energy loss per unit distance ($\langle\frac{dE}{dx}\rangle$) and longer time-of-flight (TOF) measurements than SM particles, which generally have $\beta > \num{0.9}$. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

    These four approaches cover a wide range of LLP lifetimes that would be missed in traditional searches. The next section will look  at a recent example of each analysis type from either ATLAS or CMS.

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    LLPs that live longer than most SM particles but decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

@@

\subsection{Disappearing Tracks}
    \fxnote{cite CONF-2017-017}
    \fxnote{nice disappearing track cartoon}
    \fxnote{nice cartoon feynman diagrams of signal processes}
    Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate) that eventaully decays to the neutralino LSP and, commonly, a pion. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be constructed, which results in a disappearing track.

    Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 datataking and CMS before 2017 datataking should increase sensitivity to decays at shorter radii \fxnote{I haven't mentioned previous iterations when talking about other analyses}. The ALTAS search presented here is sensitive to LLP lifetimes of approximated \SI{10}{\pico\s} to \SI{10}{\nano\s} \fxnote{make sure I'm talking about lifetime sensitivity in the same way across analyses}.

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach. The ISR jet in process (1) gives the gauginos something to recoil off of so the MET signal isn't lost \fxnote{My interpretation. Run it by someone who knows stuff. Maybe just leave this out but be prepared to explain it?}.

    Disappearing track candidates, called pixel tracklets, are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The pixel tracklets are required to pass several isolation, pT, and quality requirements such as number of pixel layer hits. Finally, the disappearing criteria is enforced by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one pixel tracklet, no electron or muon candidates (to suppress ttbar W/Z+j processes), and the MET and jet requirements associated with either of the two production modes described above.

    Pixel tracklets can be faked by random combinations of unrealted hits from nearby particles or by pixel hits that don't have the proper hits from the next tracking layer associated to them, either due to a hadron suffering a hard scatter or a lepton radiating a hard photon \fxnote{lots of words}. These backgrounds come primarily from ttbar and W+jets processes, and estimates for each type of background are obtained from data. \fxnote{I initially wrote a whole big thing about how each bg was estimated, but I'm thinking it was overkill}

    No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass gluon mass plane for the QCD production process \fxnote{hopefully will be way clearer after texing in sparticle symbols}. \fxnote{Either include plots or say numbers}

\subsection{Stopped Particles}
    \fxnote{Add calorimeter search?}  
    \fxnote{cite CMS PAS EXO-17-004}
    \fxnote{mention how stopped particles complement HSCP search?}
    \fxnote{too redundant with overview?}
    With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

    This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as Split SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos could form R-hadrons and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime.

    This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmic muons, beam halo\fxnote{explain}, and muon system noise. Beam halo and muon system noise are easily removed by the selection criteria, but distinguishing signal from cosmic muons requires a more nuanced approach. After generating signal MC and modeling cosmic muons from data obtained in dedicated cosmic runs, the authors found that cosmic muons and signal could be differentiated using time-of-flight information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres.\fxnote{Get cartoon from Juliette}

    In addition to the above criteria, the authors developed a new muon reconstruction algorithm, called displaced standalone (DSA), that uses only muon system hits and does not constrain the location of the primary vertex in any way. The full selection requires exactly one good DSA track in each hemisphere with pT greater than \SI{50}{\giga\electronvolt}, no reconstructed vertices, and several criteria related to the quality of the reconstructed track and timing measurement. Finally, restrictions on the direction of muon momentum (incoming vs outgoing), relative time measurements, and maximum number of CSC hits (0) minimize background from beam halo and cosmics.

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from \SI{100}{\nano\s} to \SI{1}{\micro\s}. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

\subsection{Heavy Stable Charged Particles}
    \fxnote{cite CMS EXO7-010}
    \fxnote{nice dE/dx plot (fig 1)}
    Particles that live for more than a couple ns \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss.

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes $\beta \approx \num{1}$. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF}

    The trigger requires all events to have either a high-pT muon or large MET. The high-pT muon trigger is more efficient for all benchmark models except the case where the R-hadron becomes neutral before reaching the MS. In the case where the R-hadron does not leave a track in the MS, the large MET trigger should recover some events.

    The authors estimate the signal-region background using an ABCD method in which events are binned in a 2-D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region \fxnote{Do I need to explain this? If I do, I should be more clear}. Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

    No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed limits and theoretical predictions, the authors also extract mass limits for the gluinos, stops, and staus in various models. \fxnote{Include plots - table is too big and too many numbers to list}

@@

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the Higgsino\fxnote{careful, there could be multiple Higgsinos. Do I mean 'lightest Higgsino'?}, which affects mH at tree level; the stop\fxnote{mention that stop is more important than other sfermions b/c large yukawa coupling? I'll have to clarify that somewhere}, which affects mH at one loop; and the gluino, which affets mStop at one loop and therefore mH at two, must all be light enough to avoid excessive fine tuning.\fxnote{diagrams?} Furthermore, the stop and gluino masses are correlated --- a large difference between the two would itself be unnatural.

    Natural SUSY scenarios, then, are most likely to be constrained by limits on the stop, gluino, and Higgsino masses. The Higgsino, however, does not carry color charge and is far less likely to be produced at the LHC,\fxnote{things get wordy and repetative after this} which means that stop and gluino limits will be the most important experimental constraints on natural SUSY. Demanding naturalness also constrains the messenger scale of SUSY breaking\fxnote{Make sure I understand this. Messenger scale != breaking scale?}. Corrections to the Higgs mass from stops and gluinos depend of the messenger scale\fxnote{reference eqs}, so natural SUSY generally prefers a lower scale.\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

@@

    Despite strong contraints, SUSY remains one of the least-tuned options for BSM naturalness\fxnote{Make sure I can explain this and make sure it isn't plagarized from Cornering}.\fxnote{Good place for Feng "how much tuning is acceptable? discussion}

@@

>>>>

\title{Naturalness and Long-Lived Beyond the Standard Model Particles}
\author{Bryan Cardwell}

@@

\section{Supersymmetry}
    \fxnote{seems like I should mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino). Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), split\fxnote{capitalize} SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker $\pT$ resolution declines \fxnote{quantify or show std plot?}.

@@

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    \fxnote{steal jamie's cartoon?}  
    LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying.

    Displaced vertex and disappearing track analyses both look for\fxnote{again, analysts look, not analyses} particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles within the inner volume \fxnote{specify} of the detector, leaving tracks that point back to a vertex some measurable distance from the nominal interaction point (IP). Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover a wide range of lifetimes and topologies for LLPs that decay on detector timescales\fxnote{dangerous as hell. Be really good at giving a quantitative explanation of this sentence}.

    Slow and stopped particle searches, on the other hand, are sensitive to detector-stable\fxnote{define?} LLPs that are far more massive\fxnote{be able to quantify} than any stable SM particle. The lower velocity of these high-mass particles shows up as increased ionization energy loss per unit distance ($\langle\frac{dE}{dx}\rangle$) or longer time-of-flight (TOF) measurements than SM particles, which generally have $\beta > \num{0.9}$\fxnote{cite something}. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy\fxnote{how long and how small?}. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings.

    These\fxnote{I've been saying `these' a lot recently} four approaches cover a wide\fxnote{how wide} range of LLP lifetimes that would be missed in traditional searches. The next section will look  at a recent example of each analysis type from either ATLAS or CMS.

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    \fxnote{Specify dataset}
    LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}\fxnote{Is it still a DV if it decays in active material somewhere?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

@@

\subsection{Disappearing Tracks}
    \fxnote{cite CONF-2017-017}
    \fxnote{nice disappearing track cartoon}
    \fxnote{nice cartoon feynman diagrams of signal processes}
    \fxnote{Specify dataset}
    Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track.

    Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 datataking and CMS before 2017 datataking should increase sensitivity to decays at shorter radii \fxnote{I haven't mentioned previous iterations when talking about other analyses}. The ALTAS search presented here is sensitive to LLP lifetimes of approximately \SI{10}{\pico\s} to \SI{10}{\nano\s} \fxnote{make sure I'm talking about lifetime sensitivity in the same way across analyses}.

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET\fxnote{add diagrams}. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable\fxnote{why?}, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach.

    Disappearing track candidates, called pixel tracklets, are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The pixel tracklets are required to pass several isolation, $\pT$\fxnote{has $\pT$ been defined?}, and quality requirements such as number of pixel layer hits. Finally, the disappearing tracks are selected by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one pixel tracklet, no electron or muon candidates (to suppress $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes\fxnote{why? be able to explain}), and the $\MET$ and jet requirements associated with either of the two production modes described above\fxnote{be able to explain} .

    Pixel tracklets can be faked by random combinations of unrealted hits from nearby particles or by tracks whose pixel hits and next tracking layer hits are not properly associated, either due to a hadron suffering a hard scatter or a lepton radiating a hard photon. These backgrounds come primarily from $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes, and estimates for each type of background are obtained from data. \fxnote{I initially wrote a whole big thing about how each bg was estimated, but I'm thinking it was overkill}

    No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass-gluon mass plane for the QCD production process \fxnote{hopefully will be way clearer after texing in sparticle symbols}. \fxnote{Either include plots or say numbers}

\subsection{Stopped Particles}
    \fxnote{Add calorimeter search?}  
    \fxnote{Specify dataset}
    \fxnote{cite CMS PAS EXO-17-004}
    \fxnote{mention how stopped particles complement HSCP search?}
    \fxnote{too redundant with overview?}
    With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

    This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as split\fxnote{capitalize?} SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos form R-hadrons\fxnote{have these been defined?} and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime.

    This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmic muons, beam halo\fxnote{explain}, and muon system noise. Beam halo and muon system noise are easily removed by selection criteria, but distinguishing signal from cosmic muons requires a more nuanced approach. After generating signal MC\fxnote{defined?} and modeling cosmic muons from data obtained in dedicated cosmic runs, the authors found that cosmic muons and signal could be differentiated using TOF information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres.\fxnote{Get cartoon from Juliette}

    In addition to the above criteria, the authors developed a new muon reconstruction algorithm, called displaced standalone (DSA), that uses only muon system hits and does not constrain the location of the primary vertex in any way. The full selection requires exactly one good DSA track in each hemisphere with $\pT$ greater than \SI{50}{\giga\electronvolt}, no reconstructed vertices\fxnote{why?}, and several criteria related to the quality of the reconstructed track and timing measurement. Finally, restrictions on the direction of muon momentum (incoming vs outgoing), relative time measurements, and maximum number of CSC hits (0) minimize background from beam halo and cosmic muons\fxnote{be able to explain how}.

    After estimating the background by extrapolating data from the cosmic-enriched\fxnote{enriched how? be able to explain} region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculate the expected background for LLP lifetimes ranging from \SI{100}{\nano\s} to \SI{1}{\micro\s}. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

\subsection{Heavy Stable Charged Particles}
    \fxnote{cite CMS EXO7-010}
    \fxnote{Specify dataset}
    \fxnote{nice dE/dx plot (fig 1)}
    Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss.

    HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge\fxnote{insert feynman diagrams}. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY\fxnote{why?}.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes $\beta \approx \num{1}$. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF?}

    The trigger requires all events to have either a high-$\pT$ muon or large $\MET$. The high-$\pT$ muon trigger is more efficient for all benchmark models except the case where the R-hadron becomes neutral before reaching the MS. In the case where the R-hadron does not leave a track in the MS, the large MET trigger should recover some events.\fxnote{Clunky wording}

    The authors estimate the signal-region background using an ABCD method in which events are binned in a 2-D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region \fxnote{Do I need to explain this? I glossed over the fact that they used both an ABCD and an ABCDEFGHI}. Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

    No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed limits and theoretical predictions, the authors also extract mass limits for the gluinos, stops, and staus in various models. \fxnote{Include plots}

@@

    Interpreting current limits in the context of naturalness requires some set of criteria that determine if a model is natural. No well-defined, universally-agreed-upon criteria exist, but certain themes recur in many discussions of natural model building\fxnote{better be able to confidently cite several sources}. The following discussion lays out a simplistic set of criteria for natural SUSY models and attempts to assess the status of naturalness in light of current experimental limits\fxnote{Did I just steal Chris's langauge?}.

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the Higgsino\fxnote{careful, there could be multiple Higgsinos. Do I mean 'lightest Higgsino'?}, which corrects $m_H$ at tree level; the stop\fxnote{mention that stop is more important than other sfermions b/c large yukawa coupling? I'll have to clarify that somewhere}, which corrects $m_H$ at one loop; and the gluino, which corrects mStop at one loop and therefore $m_H$ at two, must all be light enough to avoid excessive fine tuning.\fxnote{Need diagrams and correction terms for sure} Furthermore, the stop and gluino masses are correlated --- a large difference between the two would itself be unnatural.

    Natural SUSY scenarios, then, are most likely to be constrained by limits on the stop, gluino, and Higgsino masses. The Higgsino, however, does not carry color charge and is far less likely\fxnote{be able to quantify} to be produced at the LHC, leaving stop and gluino mass limits as the most important contraints on natural SUSY models. As shown in \fxnote{reference correction terms}, the size of stop and gluino corrections also depends on the messenger scale of SUSY breaking.\fxnote{Make sure I understand this. Messenger scale != breaking scale?}\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

    Two qualitative decisions must be made when analyzing the state of naturalness: (1) How best to define the degree of fine tuning in a particular model and (2) what degree of fine tuning is acceptable? A common, but not universal, definition for (1) is \fxnote{insert Delta eq.}, where $\alpha_i$ is a particular parameter and $\Delta_i$ is the degree of fine tuning from variations in that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$, the product of $\Delta_i$, or all $\Delta_i$ added in quadrature. Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{cite something} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural. As models are constrained, physicists seem willing to accept greater degrees of fine tuning. Instead of asking, ``what degree of fine tuning should be considered natural?", Feng\fxnote{cite} proposes physicists ask themselves, ``If SUSY were discovered with such-and-such degree of fine tuning, would I consider the naturalness problem solved and move on to different research questions?" Anecdotally, Feng reports that acceptable levels of fine tuning tend to increase when the question is posed in this way.\fxnote{break into two paragraphs?}

    Figure \fxnote{reference atlas SUSY limits} summarizes limits on SUSY models from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.

@@

    Despite strong contraints, SUSY remains one of the least-tuned options for BSM naturalness\fxnote{Make sure I can explain this and make sure it isn't plagarized from Cornering}.\fxnote{Good place for Feng ``how much tuning is acceptable? discussion}

@@

\fxnote{When should symbols be italicized and when should they be mathrm'd?}
